The Spatial and Temporal Domains of Modern Ecology

### 2017-05-11 paper/main.tex
Revision through new figure 3, and updates to text. Culling out old code that does not appear or is used in analysis (e.g. normal bootstrapping code, polygon version of plots)

\author
{Lyndon Estes$^{\ast1, 2}$, Labeeb Ahmed$^{3}$, Kelly Caylor$^{2}$, Jason Chang$^{3}$, \\
Jonathan J. Choi$^{4}$, Erle Ellis$^{3}$, Paul R. Elsen$^{1}$, and Tim Treuer$^{4}$ \\
\\
\normalsize{$^{1}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{2}$Civil and Environmental Engineering, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{4}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  lestes@princeton.edu.}
}
% Include the date command, but leave its argument blank.

@@

%\begin{sciabstract}
\textbf{Observational scale shapes our understanding of ecosystems \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}. Ecological patterns emerge at temporal and spatial scales that may be coarser or finer than the processes that shape them, thus investigation across multiple scales is essential for understanding ecological phenomena \cite{levin_problem_1992}. Given the centrality of scale to ecology, and the growing ability to observe ecological phenomena beyond a field perspective, it is important to understand the scales at which modern ecological research is conducted. To reveal these, we conducted a meta-analysis of recent (2004-2014) ecological research, in which we estimated the spatial resolution and extent, and temporal interval and duration, of reported ecological observations. Here we show that ecology remains a largely field-based discipline that makes observations within generally narrow spatial and temporal domains, despite the well-established literature on the importance of scale \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}. Ecologists still make limited use of remote sensing (6\% of observations), which is necessary for observing larger-scale patterns and processes \cite{turner_remote_2003,pettorelli_satellite_2014}, but make slightly greater use of technologies that capture high frequency or temporally continuous point-based measurements (12\% of observations). Resolution, extent, and duration were not clearly reported in most reviewed studies. This indicates inattentiveness to scale, which, together with the narrow spatio-temporal domains of observations, suggests a limited ability to generalize the knowledge gained in many ecological studies, which is critical for developing an ability to predict ecological responses in an era of increasing global change \cite{levin_problem_1992}. An emerging wave of observing and analytical technologies (e.g. nano-satellites, unmanned systems, open image archives, and high-power processing platforms) now makes it dramatically easier and cheaper to collect high-quality, ecologically meaningful measurements at scales inaccessible to field scientists. Combined with a greater attention to scale in research design, these advances can facilitate the spread of ecological research to sparsely observed scales, while improving the transferability of ecological knowledge.}

@@

\linenumbers
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is the \emph{sine qua non} for understanding ecological phenomena \cite{levin_problem_1992}. Awareness of the importance of scale has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad range of scales \cite{schneider_rise_2001,chave_problem_2013}.

Given the centrality of scale to understanding ecology, and the growing ability to study ecological phenomena across a broad range of spatial and temporal scales, it is important to assess the scales at which ecological research is presently being conducted. To gain insight into this question, we quantified the spatial and temporal domains of empirical observations that were reported in a representative sample of studies published between 2004-2014 in the top 30 ecological journals (by impact factor). Empirical observations provide the necessary means for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, thus it stands to reason that the temporal and spatial distributions of ecological observations may be indicative of modern ecology's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

We characterized observational domains along two key spatial dimensions, resolution (grain) and extent, and their temporal corollaries, sampling interval and total temporal duration. We calculated the actual, as opposed to representative (SI), scales for these four dimensions from 371 discrete observations (defined here as data collected from non-experimentally manipulated, or ``natural'' \cite{tilman_ecological_1989}, systems) reported within a 140 paper subset of 346 randomly selected articles (from 42,918 total). An additional 56 papers that were cited as the source of observations in the selected papers were also reviewed. To account for uncertainty in the estimation of observational dimensions due to 1) unclear methodological description in the reviewed papers, and 2) observer interpretation, we conducted a resampling analysis (n=1000) in which scale values were randomly perturbed within the bounds of estimated inter-observer variation (SI). We constructed histograms for each dimension from the mean of the perturbed ensembles, and estimated 95\% confidence intervals for each histogram bin (Fig. 1). We constructed  kernel density estimates from the full resampled ensemble in order to assess observational distributions within different juxtapositions of the four space-time dimensions (Fig. 2).

In terms of spatial resolution (the two-dimensional space in which all measurable features of a natural object(s) were recorded, as opposed to sub-sampled), the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The total spatial extent (the number of sampled sites multiplied by the spatial resolution) of 86\% of observations was $<$10 ha, while 36\% covered less than 1 m$^2$ (Fig. 1B). Only 14\% covered an extent $\geq$10 ha, with just 4\% spanning areas $\geq$1 million ha.

>>>>

\author
{Lyndon Estes$^{\ast1, 2, 3}$, Paul R. Elsen$^{1, 4}$, Tim Treuer$^{5}$, Labeeb Ahmed$^{6}$, \\
Kelly Caylor$^{1, 7}$, Jason Chang$^{6}$, Jonathan J. Choi$^{5}$, and Erle Ellis$^{6}$ \\
\\
\normalsize{$^{1}$Woodrow Wilson School, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{2}$Civil and Environmental Engineering, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{3}$Graduate School of Geography, Clark University, Worcester MA 01610, USA}\\
\normalsize{$^{4}$Department of Environmental Science, Policy, and Management,}\\
\normalsize{University of California Berkeley, Berkeley, CA 94720, USA}\\
\normalsize{$^{5}$Ecology and Evolutionary Biology, Princeton University, Princeton, NJ 08544, USA}\\
\normalsize{$^{6}$Geography and Environmental Systems, University of Maryland Baltimore County,}\\ 
\normalsize{Baltimore, MD 21250, USA}\\
\normalsize{$^{7}$Earth Research Institute, University of California Santa Barbara, Santa Barbara, CA 93106, USA}\\
\\
\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  LEstes@clarku.edu.}}
% Include the date command, but leave its argument blank.

@@

%\begin{sciabstract}
\section*{Abstract}
%Ecologists have long known that the observational scales play a critical role in shaping our understanding of ecosystems, and that observations spanning multiple scales of space and time are needed to effectively characterize ecological phenomena. 
\textbf{In order to properly understand ecological phenomena, it is necessary to quantify their behavior over the range of spatial and temporal scales at which they manifest. Ecology has been concerned with this need since the early 1990s, and the ability to collect multi-scaled ecological observations has grown rapidly since then.  Characterizing the spatial and temporal domains of modern ecological observations can therefore provide important insight into the field's progress in understanding towards a more comprehensive understanding of ecosystem behaviour. To characterize these domains, we conducted a meta-analysis of recent (2004-2014) ecological studies, in which we quantified four primary dimensions of their reported observations: plot resolution, sampling interval, effective duration (time between start and end of temporal replicates), and effective extent (area enclosed by spatial replicates). We also estimated the \emph{actual} extent and duration, which respectively represent the summed area and time covered by spatial and temporal replicates. \textcolor{blue}{Replace this text with more specific summary of observed scales:} \textcolor{red}{Here we show that ecology remains a largely field-based discipline that makes observations within generally narrow spatial and temporal domains, despite the well-established literature on the importance of scale \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}}}. % Ecologists still make limited use of remote sensing (6\% of observations), which is necessary for observing larger-scale patterns and processes \cite{turner_remote_2003,pettorelli_satellite_2014}, but make slightly greater use of technologies that capture high frequency or temporally continuous point-based measurements (12\% of observations). Resolution, extent, and duration were not clearly reported in most reviewed studies. This indicates inattentiveness to scale, which, together with the narrow spatio-temporal domains of observations, suggests a limited ability to generalize the knowledge gained in many ecological studies, which is critical for developing an ability to predict ecological responses in an era of increasing global change \cite{levin_problem_1992}. An emerging wave of observing and analytical technologies (e.g. nano-satellites, unmanned systems, open image archives, and high-power processing platforms) now makes it dramatically easier and cheaper to collect high-quality, ecologically meaningful measurements at scales inaccessible to field scientists. Combined with a greater attention to scale in research design, these advances can facilitate the spread of ecological research to sparsely observed scales, while improving the transferability of ecological knowledge.}

@@

\linenumbers
\vspace{10pt}
The scales at which ecosystems are observed plays a critical role in shaping our understanding of their structure and function \cite{levin_problem_1992,chave_problem_2013,wiens_spatial_1989}.  Ecological patterns emerge from temporal and spatial domains that may be coarser or finer than the processes that shape them, which means that investigation across multiple scales is the \emph{sine qua non} for understanding ecological phenomena \cite{levin_problem_1992}. This awareness has grown rapidly since the 1980s, accelerated by the need to understand how changes in the global climate, ocean, and land systems are affecting everything from individual populations \cite{tingley_push_2012} to entire biomes \cite{xiao_photosynthetic_2004}, while technological advances in areas such as remote sensing and genetics are making it ever-easier to quantify ecological features across a broad range of scales \cite{schneider_rise_2001, chave_problem_2013}.

Given this awareness of the centrality of scale to understanding ecology, and the growing ability to study ecological phenomena across a broad range of spatial and temporal scales, it is important to assess the scales at which current ecological research is conducted. To gain insight into this question, we quantified the spatial and temporal domains of empirical observations that were reported in a representative sample of studies published between 2004-2014 in the top 30 ecological journals (by impact factor). Empirical observations provide the necessary means for developing and testing the models that explain why ecological patterns vary in time and space \cite{levin_problem_1992, tilman_ecological_1989}, thus it stands to reason that the temporal and spatial distributions of ecological observations may be indicative of modern ecology's progress towards achieving a holistic, predictive understanding of ecosystems \cite{chave_problem_2013,levin_problem_1992}.

We characterized observational domains along two key spatial dimensions, resolution (grain) and extent, and their temporal corollaries, interval and duration (Table 1).  Here resolution is the area of an individual spatial replicate, or the two-dimensional space in which all measurable features of a natural object(s) were recorded (as opposed to sub-sampled), while extent is the area enclosed by the outer-most spatial replicates, or, if the system or habitat being sampled was distinct from its surrounding matrix (e.g. forest patches in grassland habitats), the summed area of sampled patches (see SI for full definition). Interval refers to the average time elapsed between individual temporal replicates, and duration is the time elapsed between the first and last temporal replicates, or, in the case of temporally unreplicated observations, the estimated time spent collecting the observation (SI). We also calculated two additional metrics, the integrated area of spatial replicates (i.e. resolution multiplied by number of replicates) and the summed observational time of all temporal replicates. We estimated these additional dimensions to evaluate the degree to which the actual scales of ecological observations differ from those they ostensibly represent, and therefore refer to them as the \emph{actual} extent and \emph{actual} duration.

\begin{table}[htp]
\caption{The dimensions of ecological observations estimated in this meta-analysis.}
\begin{center}
\begin{tabular}{ll}
\hline
\textbf{Dimension} & \textbf{Description} \\
\hline
Resolution & Area (m$^2$) of an individual spatial replicate (plot) \\
Extent & Area (ha) encompassed by all spatial replicates \\
Actual extent & Summed area (ha) of all spatial replicates\\
Interval & Time elapsed (days) between successive temporal replicates \\
Duration & Time elapsed (days) between first and last temporal replicates\\
Actual duration & Summed observational time (days) of all temporal replicates\\
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%

We calculated these dimensions from 379 discrete observations (defined here as data collected from non-experimentally manipulated, or ``natural'' \cite{tilman_ecological_1989}, systems) reported within a 134 paper subset of 348 randomly selected articles (from 42,918 total). An additional 62 papers that were cited as the source of observations in the selected papers were also reviewed.

To account for uncertainty in the estimation of observational dimensions due to 1) unclear methodological description in the reviewed papers, and 2) observer interpretation, we conducted a resampling analysis (n=1000) in which scale values were randomly perturbed within the bounds of estimated inter-observer variation (SI). We constructed histograms for each dimension from the mean of the perturbed ensembles, and estimated 95\% confidence intervals for each histogram bin (Fig. 1). We constructed  kernel density estimates from the full resampled ensemble in order to assess observational distributions within different juxtapositions of the four space-time dimensions (Fig. 2).

In terms of resolution, the majority (67\%) of observations were collected in plots of $<$1 m$^2$ resolution, 24\% were collected within plots of 1 m$^2$ up to 1 ha, and the remaining 9\% in plots of $\geq$1 ha (Fig. 1A). The total spatial extent (the number of sampled sites multiplied by the spatial resolution) of 86\% of observations was $<$10 ha, while 36\% covered less than 1 m$^2$ (Fig. 1B). Only 14\% covered an extent $\geq$10 ha, with just 4\% spanning areas $\geq$1 million ha.
